UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-09-07 ms/main.tex
reword pgh

By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure. Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP. Building the global structure in t-SNE requires tuning the \textit{perplexity} parameter, which in practice will need to be different for varying cluster sizes\cite{wattenberg2016how}, while UMAP merges discrete simplicial sets into a coherent global structure. Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}.

@@

\subsubsection*{Caveats}
In contrast to PCA, UMAP has more adjustable parameters. Changing the PC cutoff, minimum distance, and number of neighbours can change characteristics of the visualizations. Using a minimal number of neighbours (e.g. 5) can result in the formation of disjoint clusters comprised of related family members (fig~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail. We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally perform well across datasets.

@@

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in fig~\ref{fig:supp_megamontage_pc10_50}. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual space on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figs, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

>>>>

%By patching together many local genetic neighbourhoods, UMAP visualizations preserve aspects of both local and global structure. Methods such as t-SNE and LargeVis\cite{tang2016visualizing} are similarly motivated, with key differences in how they measure neighbours and which cost functions are used in optimizing low-dimensional embeddings: pairwise Gaussian and t-distributed approximations and the KL divergence in t-SNE; approximate nearest neighbours and a likelihood function in LargeVis; and fuzzy simplicial set membership and the cross-entropy in UMAP. Further practical and theoretical comparisons can be found in~\cite{2018arXivUMAP}. Hierarchical clustering of networks has also been successful in detecting fine-scale population structure, though it was instead motivated by partitioning IBD data into subnetworks to then find stable subclusters\cite{han2017clustering}; in practice, one could carry this out by detecting clusters in data that has been pre-processed using UMAP as laid out in \cite{howtoUMAP}.

Methods such as UMAP and t-SNE focus on preserving local distances to reveal fine-scale structure in populations, and in the process may preserve aspects of global structure as well. In contrast, PCA preserves long range distances but hides finer-scale details. Hierarchical clustering of networks has also successfully detected fine-scale population structure using identity-by-descent similarity by attempting to preserve relations between global networks and smaller local ones\cite{han2017clustering}. We speculate that the addition of weak constraints favouring the preservation of longer distances in UMAP-like approaches has the potential of preserving the desirable local properties while encouraging more intuitive positioning of clusters on a global scale.

@@

\subsubsection*{Caveats}
In contrast to PCA, UMAP has more adjustable parameters. Changing the PC cutoff, minimum distance, and number of neighbours can change characteristics of the visualizations. Using a minimal number of neighbours (e.g. 5 rather than the default 15) can result in the formation of disjoint clusters comprised of related family members (fig~\ref{fig:supp_1kgp_families}), and using a low minimum distance (e.g. 0.001 rather than the default 0.1) can result in clusters becoming more compact, losing visual detail. We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP generally perform well across datasets.

@@

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in fig~\ref{fig:supp_megamontage_pc10_50}. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual space on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.
